---
title: "Univariate Distributions as `xDens` Objects"
author: "Martin Lysy, Jonathan Ramkissoon"
date: "`r Sys.Date()`"
output:
  html_vignette:
    toc: true
bibliography: references.bib
csl: taylor-and-francis-harvard-x.csl
link-citations: true
vignette: >
  %\VignetteIndexEntry{xDensity Objects}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, eval = FALSE, echo = FALSE}

# R code to render vignette
rmarkdown::render("xDensity-quicktut.Rmd")

```

<!-- Latex commands -->
\newcommand{\bm}[1]{\boldsymbol{#1}}
\newcommand{\tx}[1]{\mathrm{#1}}
\newcommand{\rv}[3][1]{#2_{#1},\ldots,#2_{#3}}
\newcommand{\XX}{{\bm X}}
\newcommand{\xx}{{\bm x}}
\newcommand{\uu}{{\bm u}}
\newcommand{\iid}{\stackrel{\mathrm{iid}}{\sim}}

## Introduction

The **`GaussCop`** package proposes a simple, unified framework to provide the basic `d/p/q/r` functions for arbitrary continuous one-dimensional distributions.  This can be useful for working with various Copula models (including the Gaussian Copula). 

### `xDens` Objects

Let $X$ denote an unbounded, continuous random variable with PDF $X \sim f(x)$.  An `xDens` object essentially consists of the following elements:

 - PDF values on an evenly-spaced grid.
 - The mean and standard deviation of a Normal distribution to be used outside the grid.
 
 Based on this representation, the `d`/`p`/`q`/`r` functions are implemented as follows:
 
 - *PDF* (`d`): Step function inside the grid, Normal outside the grid.
 - *CDF* (`p`): Piecewise linear inside the grid, Normal outside the grid.
 - *Inverse-CDF* (`q`): Exact numerical inversion of the CDF.
 - *Sampling* (`r`): Inverse-CDF method.  That is, if $F(x)$ is a CDF and $U \sim \tx{Unif}(0,1)$, then $X = F^{-1}(U)$ has CDF $F(x)$.
 
### Basic Usage

```{r}


```

`xDens` objects  can be constructed either from iid samples $\rv X n \iid f(x)$ or from a functional representation of $f(x)$ itself.  
<!-- are created from a non-parametric estimate of $f(x)$ obtained from a random sample $\rv X n \iid f(x)$.   -->
For example, consider the following `xDens` representation based on a random sample from the log-Exponential distribution,
$$
\log(X) \sim \tx{Expo}(1) \qquad \iff \qquad f(x) = \exp(-e^x + x).
$$
```{r}
require(GaussCop) # load package

# simulate data
nsamples <- 1e5
X <- log(rexp(nsamples)) # iid samples from the log-Exponential

xDens <- kernelXD(X) # basic xDens constructor
names(xDens)
```
The elements of `xDens` are:

- `ndens`, `xrng`: a compact representation of the $x$-value grid, i.e., the number of grid points and the range of the grid.
- `ypdf`, `ylpdf`, `ycdf`:  the PDF, log-PDF, and CDF evaluated at the `ndens` grid points.  While the latter two elements can be reconstructed from the first, they are stored here in order to speed up subsequent calculations.
- `mean`, `sd`: The mean and variance of the Normal distribution used outside the range of the grid.  By default, these values are chosen by matching the Normal PDF to the `ypdf` values at then endpoints of the grid; details of the calculation can be found in the [Appendix](#appendix) .
<!-- contains the density at the grid midpoint, and `mean` and `sd` contain the PDF of the Normal outside the grid.  In addition, `ylpdf` and `ycdf` contain the log-density and CDF on the grid, which are precomputed to speed up subsequent calculations. -->

The `d/p/q/r` functions for `xDensity` objects are invoked by e.g., `dXD(x, xDens = xDens, log = TRUE)` as seen in the [Appendix](#apptest) **R** code for the test function `xdens_test`.
<!-- First we create a generic function to compare a true distribution to its `xDens` representation: -->
```{r xdens_test, echo = FALSE}
# check that xDens d/p/q/r functions match the true distribution,
# by plotting true PDF/CDF vs xDens representation.
# xDens: xDens object.
# dtrue/ptrue: single-variable function of the true pdf/cdf.
# dname: name of true distribution (for legend)
# dlpos, plpos: legend position for pdf/cdf plot.
xdens_test <- function(xDens, dtrue, ptrue, dname, dlpos, plpos) {
  op <- par(no.readonly = TRUE)
  par(mfrow = c(1,2), mar = c(4,4,2,.5)+.1)
  # pdf and sampling check
  nsamples <- 1e5
  Xsim <- rXD(nsamples, xDens = xDens) # xDens: random sampling
  # extend xDens range to show endpoint matching
  xlim <- xDens$xrng
  xlim <- (xlim - mean(xlim)) * 1.10 + mean(xlim)
  hist(Xsim, breaks = 100, freq = FALSE, # xDens random sample
       xlab = "x", main = "PDF", xlim = xlim)
  curve(dXD(x, xDens = xDens), add = TRUE, col = "red") # xDens PDF
  curve(dtrue, add = TRUE, col = "blue") # true PDF
  abline(v = xDens$xrng, lty = 2) # grid endpoints
  legend(dlpos, parse(text = c(dname, "dXD", "rXD", "xrng")),
         pch = c(22,22,22,NA), pt.cex = 1.5,
         pt.bg = c("blue", "red", "white", "black"),
         lty = c(NA, NA, NA, 2))
  # cdf check
  curve(pXD(x, xDens), # xDens CDF
        from = xlim[1], to = xlim[2], col = "red",
        xlab = "x", main = "CDF", ylab = "Cumulative Probability")
  curve(ptrue, add = TRUE, col = "blue") # true CDF
  abline(v = xDens$xrng, lty = 2) # grid endpoints
  legend(plpos, parse(text = c(dname, "pXD", "xrng")),
         pch = c(22,22,NA), pt.cex = 1.5,
         pt.bg = c("blue", "red", "black"),
         lty = c(NA, NA, 2))
  invisible(par(op))
}
```
That is, consider the following graphical check of an `xDensity` approximation to the log-Exponential distribution $\log(X) \sim \tx{Expo}(1)$:

1. Check that histogram of `rXD` sample matches `dXD` and the true PDF (recall that `rXD` calls `qXD` on `runif`).
2. Check that `pXD` matches true CDF.

These graphical tests are implemented by `xdens_test` as illustrated below:
```{r, fig.width = 10, fig.height = 4, out.width = "97%", fig.cap = "Accuracy test for `xDens` approximation of `d/p/q/r` functions."}
# check that xDens d/p/q/r functions match the true distribution,
# by plotting true PDF/CDF vs xDens representation.

dlexp <- function(x) exp(-exp(x) + x) # true PDF
plexp <- function(x) pexp(exp(x)) # true CDF

xdens_test(xDens, # xDensity approximation 
           dtrue = dlexp, # true PDF 
           ptrue = plexp, # true CDF
           dname = expression(log(X)%~%"Expo"*(1)), # name of true distribution (for legend)
           dlpos = "topleft", plpos = "topleft") # legend position for pdf/cdf plot

```


## Constructors

Power Transform:
$$z = \begin{array}{rl} \frac{((x + \alpha)^\lambda - 1)}{(\lambda C^{\lambda-1})} & \lambda \neq 0 \\ C \log(x + \alpha) & \lambda = 0, \end{array}$$

`xDens` objects can be constructed using three methods: two involving density estimation of random sample from $f(x)$, and one from a functional approximation to $f(x)$, e.g., using a parametric family $f(x \mid \hat {\bm \theta})$, where $\hat{\bm \theta}$ is the parameter MLE.  
<!-- These methods are discussed and compared using the true distribution $X \sim \chi^2_{(4)}$. -->

<!-- approximate densities using one of 3 methods: kernel density estimation, Gram-Charlier expansion or by using known a known sample-density matrix. These are discussed and compared using the $\chi^2_{(5)}$ distribution. An endpoint matching technique is used to describe the tails of the approximated densities with a Normal distribution. More about this can be found in [Appendix A: Endpoint Matching]. -->


### Kernel Density Estimation

The `kernelXD` function uses the `density` function in **R** to fit a Gaussian kernel to a random sample from $f(x)$. It is the most flexible constructor, however it also requires the most datapoints for stable estimation. Below is a comparison with $M$ iid samples from a $\chi^2_{(4)}$ distribution, using $M = 1000$ and $M = 10000$ data points.

```{r, fig.width = 10, fig.height = 4, out.width = "97%", fig.show = "hold"}

# true distribution: chi^2_(4)
df_true <- 4
dtrue <- function(x) dchisq(x, df = df_true)
ptrue <- function(x) pchisq(x, df = df_true)
rtrue <- function(n) rchisq(n, df = df_true)
dname <- paste0("X%~%chi[(",df_true,")]^2")

# simulate data
nsamples <- 1e3
X <- rtrue(nsamples)
xDens <- kernelXD(X) # xDens constructor

par(oma = c(0,2,0,0)) # put sample size in outer margin
xdens_test(xDens, dtrue = dtrue, ptrue = ptrue,
           dname = dname,
           dlpos = "topright", plpos = "bottomright")
mtext(text = paste0("M = ", nsamples),
      side = 2, line = .5, outer = TRUE)

# increase sample size
nsamples <- 1e4
X <- rtrue(nsamples)
xDens <- kernelXD(X)
xdens_test(xDens, dtrue = dtrue, ptrue = ptrue,
           dname = dname,
           dlpos = "topright", plpos = "bottomright")
mtext(text = paste0("M = ", nsamples),
      side = 2, line = .5, outer = TRUE)
par(oma = c(0,0,0,0)) # remove outer margin

```

### Gram-Charlier Expansion

The `gc4XD` constructor is a moment-based density estimator, ideal for unimodal distributions to be estimated from small sample sizes.  The estimator first applies a Box-Cox transformation to the data, then fits it with a 4th order  [Gram-Charlier expansion](https://en.wikipedia.org/wiki/Edgeworth_series) [e.g., @draper.tierney72].  The example below illustrates the approximation for $M = 1000$ observations of $X \sim \chi^2_{(4)}$.

```{r, fig.width = 10, fig.height = 4, out.width = "97%"}
nsamples <- 1e3
X <- rtrue(nsamples)
xDens <- gc4XD(X) # xDens constructor
xdens_test(xDens, dtrue = dtrue, ptrue = ptrue,
           dname = dname,
           dlpos = "topright", plpos = "bottomright")
```


### Known Density Values

The `matrixXD` is used to convert functional density representations into the standardized `xDens` format. Its input is a 2-column matrix with equally-spaced $x$-grid values in the first column and corresponding density values in the second. 
<!-- It essentially appends Normally distributed tails to a grid of known PDF densities.  -->
<!-- Currently, this function does not accept zero-valued grid density endpoints.  -->
Below is an illustration using the Cauchy distribution.

```{r, fig.width = 10, fig.height = 4, out.width = "97%"}
# true distribution: Cauchy
dtrue <- function(x) dcauchy(x)
ptrue <- function(x) pcauchy(x)
dname <- expression(X%~%"Cauchy")

# matrix density representation
xgrid <- seq(from=-15, to=15, len = 500) # grid
ypdf <- dtrue(xgrid)
xDens <- matrixXD(cbind(X = xgrid, Y = ypdf)) # xDens constructor

xdens_test(xDens, dtrue = dtrue, ptrue = ptrue,
           dname = dname,
           dlpos = "topright", plpos = "bottomright")

```

## Example: The Gaussian Copula Distribution

Consider a $d$-dimensional random variable $\XX = (\rv X d)$ with unknown PDF $\XX \sim f(\xx)$.  A popular approach to statistical dependence modeling is to do so upon transforming $\XX$ to having univariate margins.  That is, if $F_i(x_i)$ is the CDF of $X_i$, then
$$
U_i = F_i^{-1}(X_i) \sim \tx{Unif}(0,1),
$$
such that the joint PDF of $\XX$ can be written as [*get density right*]
$$
f(\xx) = c(\uu), \qquad u_i = F_i^{-1}(x_i)
$$
Thus, $c(\uu)$ is the joint PDF of marginally Uniform random variables and is known as a Copula distribution.  Numerous **R** packages provide various methods for fitting and sampling from Copula distributions, e.g., [**`copula`**](https://CRAN.R-project.org/package=copula) [@hofert.etal17],  [**`VineCopula`**](https://CRAN.R-project.org/package=VineCopula) [@schepsmeier.etal17], and many others listed in the [**`Distributions`**](https://CRAN.R-project.org/view=Distributions) Task View.    

## Appendix A: Endpoint Matching {#appendix}

We wish to find the parameters $(\mu, \sigma)$ of a normal distribution whose PDF passes through the points $(x_i, y_i)$, $i = 1,2$.  Thus we wish to find $(\mu, \sigma)$ satisfying
$$
\log y_i = -\frac 1 2 \left[\frac{(x_i - \mu)^2}{\sigma^2} + \log \sigma^2 + \log 2\pi \right], \qquad i = 1,2.
$$
Subtracting one equation from the other, we obtain
$$
\log(y_1/y_2) = \frac{(x_1 - \mu)^2}{2\sigma^2} - \frac{(x_2 - \mu)^2}{2\sigma^2},
$$
such that for any value of $\sigma$ the required value of $\mu$ is
$$
\mu(\sigma) = \sigma^2\frac{\log(y_1/y_2)}{x_1 - x_2} + \frac{x_1 + x_2}{2} = b \sigma^2 + a.
$$
Substituting this back into one of the two original equations leads to the root-finding problem
$$
g(\tau) = \frac{A^2}{\tau} + B^2 \tau + \log \tau - C = 0,
$$
where $\tau = \sigma^2$ and
$$
A = x_i - a, \qquad B = b, \qquad C = 2[b(x_i-a) - \log y_i] - \log 2\pi.
$$
A solution to this equation is not guaranteed to exist.  For instance, let $x_1 = -1$, $x_2 = 1$, and $y_i = 1$.  Then a Normal distribution passing through these points must have density at least $2$.  However, in our case $y_i$ will typically be very small, in which case a root is likely to exist.  Even then, there could be two roots to the equation: one with $\mu$ between $x_1$ and $x_2$, and one with $\mu$ (way) off to one side.  We'll focus here on finding the former.

The root we seek can be calculated numerically with the (built-in) **R** function `uniroot`, which requires specification of an interval containing the root.  To obtain this interval, first notice that $g'(\tau) = -\frac{A^2}{\tau^2} + B^2 + \frac{1}{\tau}$, of which the only positive solution is
$$
\tau_L = \frac{-1 + \sqrt{1 + 4(AB)^2}}{2B^2}.
$$
Since $g(\tau) \to \infty$ for $\tau \to 0$ and $\tau \to \infty$, we conclude that $\tau_L$ is the minimum value of $g(\tau)$ on $\tau \in (0,\infty)$.  Let us now find a value $\tau_U < \tau_L$ at which $g(\tau_U) > 0$ (doing this for $\tau_U > \tau_L$ is likely to have $\mu$ outside of $(x_1, x_2)$).

In order to find a lower bound for the root, note that for $0 < \tau < A^2/2$, we have 
$$\log \tau > -\frac{A^2/2}{\tau} + D, \qquad D = 1 + \log(A^2/2).
$$
This is because the derivative of the RHS is larger than that of the LHS for $\tau < A^2/2$, thus falls to $-\infty$ monotonically faster as $\tau \to 0$.  The value of $D$ is chosen such that LHS and RHS are equal at $\tau = A^2/2$.  Substituting the inequality above into the root-finding equation, we obtain
$$
g(\tau) > \frac{A^2}{\tau} + \log \tau - C > \frac{A^2/2}{\tau} + D - C,
$$
such that $g(\tau) > 0$ for $\tau < \frac{A^2}{2(C-D)}$.

Note that when $y_1 = y_2$, the endpoint matching method breaks down and the mean and standard deviation of the approximated density is used for the Normally distributed tails.

## Appendix B: Code For Test Function `xdens_test` {#apptest}

```{r, ref.label = "xdens_test", eval = FALSE}
```

## References

<!-- - $g(\tau) \to \infty$ both as $\tau \to 0$ and $\tau \to \infty$. -->
<!-- - $g'(\tau) = -\frac{A^2}{\tau^2} + B^2 + \frac{1}{\tau}$, of which the only positive solution is -->
<!--     $$ -->
<!--     \tau = \frac{-1 + \sqrt{1 + 4(AB)^2}}{2B^2}. -->
<!-- 	$$ -->
<!--     Since $g(\tau) \to \infty$ is infinite for $\tau,1/\tau -->

<!-- Given the endpoints of the grid, $(x_1, x_2)$ and their corresponding log-densities, $(\log(y_1), \log(y_2))$, the $\mu$ and $\sigma$ of a Normal distribution for the tails is calculated as follows: -->
<!-- $$ y_i = \frac{1}{\sqrt{2\pi\sigma^2}} + e^\frac{(x_i-\mu)^2}{\sigma^2} ~~~~~~~~~~ i=1,2$$ -->
<!-- $$ \log(y_i) = -\frac{1}{2}\log(2\pi\sigma^2) - \frac{(x_i-\mu)^2}{2\sigma^2} $$ -->
<!-- Subtracting $\log(y_1) - \log(y_2)$ and denoting $\log(y_i)$ as $ly_i$ -->
<!-- $$ \frac{(x_1 - \mu)^2}{2\sigma^2} + \frac{(x_2 - \mu)^2}{2\sigma^2} = ly_1 - ly_2$$ -->
<!-- We get a linear relationship between $\mu$ and $\sigma^2$ -->
<!-- $$ \mu = \sigma^2\frac{(ly_1 - ly_2)}{x_1 - x_2} + \frac{x_1 + x_2}{2} $$ -->
<!-- For simplicity, let $a = \frac{x_1 + x_2}{2}$ and $b = \frac{(ly_1 - ly_2)}{x_1 - x_2}$, so that $\mu = b\sigma^2 + a$ -->
<!-- Substituting $\mu$ back into the log-pdf  -->
<!-- $$ ly_i = -\frac{1}{2}\log(2\pi\sigma^2) - \frac{(x_i - a - b\sigma^2)^2}{2\sigma^2} $$  -->
<!-- $$ -2ly_i = \log(\sigma^2) + \frac{(x_i - a)^2}{\sigma^2} + b^2\sigma^2 - 2(x_i - a)b + \log(2\pi) $$ -->
<!-- $$ \log(\sigma^2) + \frac{(x_i - a)^2}{\sigma^2} + b^2\sigma^2  = 2(x_i - a)b - \log(2\pi) - 2ly_i$$ -->
<!-- Letting $w = 2(x_i - a)b - \log(2\pi) - 2ly_i$, we have [[[[[[[[could $b$ be negative (yes, if $y_1 > y_2$), what to do in this situation?]]]]]]]] -->
<!-- $$ \log(\sigma^2) + \frac{(x_i - a)^2}{\sigma^2} + b^2\sigma^2 - w = 0$$ -->
<!-- $$ g(\sigma^2 ~|~ x, a, b) = \log(\sigma^2) + \frac{(x - a)^2}{\sigma^2} + b^2\sigma^2 - w $$ -->
<!-- $$ g'(\sigma^2 ~|~ x, a, b) = \frac{2}{\sigma} - \frac{2(x - a)^2}{\sigma^3} + 2b^2\sigma$$ -->
<!-- Setting $g'(\sigma^2 ~|~ x, a, b) = 0$ gives a quadratic equation in $\tau = \sigma^2$ whose roots are: -->
<!-- $$ \frac{-1 \pm \sqrt{1+4b(x-a)^2}}{2b} $$ -->
<!-- Since $\sigma^2 > 0$, the only root is $\sigma^2 = \frac{-1 + \sqrt{1+4b(x-a)^2}}{2b}$ -->



